The Effect of Oil Prices on Field Production: Evidence from the Norwegian Continental Shelf

### 2016-03-11 oil_prices_obes_blind.tex
final resubmission

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{natbib} %for citet and citep
\usepackage{syntonly}
\usepackage{esdiff} %for writing partial derivatives
\usepackage{url} %for inserting urls
\usepackage{placeins} % for limitting floats
\usepackage{comment} %for excluding figures
\usepackage{booktabs}
% \excludecomment{figure} % for printing without figures
% \syntaxonly % for quickly checking document
%set document settings

\doublespacing % from package setspacs

\title{The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf}
\author{Johannes Mauritzen\\
		Department of Economics\\
        BI Norwegian Business School\\
        Trondheim Campus\\
		Trondheim, Norway\\
        johannes.mauritzen@bi.no\\
        \url{jmaurit.github.io}\\
		}
\date{\today}

@@

\begin{abstract}
I use detailed field-level data on Norwegian offshore oil production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production. I find no significant evidence of a concurrent reaction of field production to oil prices, though a modest lagged effect is found of the magnitude of approximately 2 to 7\% for a 10 dollar per barrel increase in the real price of oil.\\
Keywords: Oil production,Oil Prices, Norway, Generalized Additive Model.
Words: 7100
\end{abstract}

@@

For most of the last century, crude oil has been the world's single most important and valuable fuel source.\footnote{Oil is the world's largest single energy source, consisting of approximately 33\% of total energy consumption in 2013 as well as the most valuable in terms of market price per unit of energy \citep{british_petroleum_statistical_2013}} Naturally, questions of how the oil price affects the world economy as well as how oil production reacts to the oil price have been fundamental topics in Economics. Recent volatility in the price of oil has made this an especially relevant topic.

Oil and gas supply, especially in difficult and expensive offshore areas, is the product of an often lengthy and complex chain of conditional investment decisions. Starting with the exploration phase, where a potential oil-producing region or "play" is explored, leases are acquired, prospects are identified and exploration wells are drilled. In the case of a significant hydrocarbon find, a field is appraised to judge size, difficulty and in turn profitability. For fields seen as sufficiently profitable, a development stage commences, where a detailed drilling and production plan is drawn up and submitted to obtain permits and regulatory approval. Operating agreements are signed and subcontractors acquired for offshore construction and well development. Finally, at the end of this multi-year process, a build-out can commence, where full production capacity is built out as fast as possible, followed by an extended depletion phase. All of this consists of what is referred to as upstream activity, with steps such as shipping, refining and trading, among others, making up an equally complex down-stream process. \footnote{\citet{lefler_deepwater_2011} provides an accessible guide to the offshore petroleum exploration and production industry.}

A large body of economic literature exists on the complex chain of investment and production decisions made in the upstream oil and gas industry. This article looks at the limited, but important topic of the effect of price on production from active offshore fields, where a significant gap exists in the literature.

While numerous studies have taken up the issue of how the price of oil affects searching for new fields as well as total oil production at both the regional and global level, few studies exist of the effects of oil prices on oil production at the field level. The studies that do use field- and well-level data, such as \citet{rao_taxation_2010}, often use data from on-shore installations.  However an increasing amount of the world's oil production comes from hard-to-reach offshore fields.  Production from challenging offshore environments is substantially different in character than from on-shore installations.

The effect of oil prices on producing fields is an important topic in understanding the mechanisms of how total oil supply reacts in response to price. The response of oil production in existing fields is especially important now as many of the major oil-producing areas, like the Norwegian Continental Shelf are mature and an increasing share of total investment will be directed at capacity enhancement in producing fields rather than new field developments. Figure \ref{investment} shows that investments in producing fields makes up a large and growing share of total investment, nearly dwarfing investments in searching, exploration and new build-outs. How production from mature fields will respond to changes in oil prices has major implications for the oil industry, the state finances of oil-producing countries and long-run oil price formation.

\begin{figure}
	\center
	\includegraphics[width=.9\textwidth]{figures/investment.png}
	\caption{The figure shows investment (millions NOK) in the oil and gas extraction industry on the Norwegian Continental Shelf. A large and growing share of total investment goes into fields that are already producing as opposed to investment in search, exploration and build-out of new fields.}
	\label{investment}
\end{figure}

The question of the effect of oil prices on production and the more general question of optimal oil extraction has spawned a large theoretical literature dating back to the seminal work of \citet{hotelling_economics_1931}. \citet{krautkraemer_nonrenewable_1998} provides a good overview. At a basic level a central idea of much of this theoretical work is that with a non-renewable resource, production is a decision that involves a significant opportunity cost: more production in the current period means less production in future periods.  Within these frameworks, prices and expectations of prices become important variables in the production decision. A simple Hotelling model suggests that a producer would immediately change their production in response to a change in oil price in order to dynamically optimize the total extraction.

But in practice, the question is not as clear cut.  Production on the Norwegian Continental Shelf - as well as most other offshore production areas - has extremely high fixed and operating costs.  Keeping spare capacity available in order to adjust to changing oil prices might simply be too expensive.  Instead, producers may find it more beneficial to use storage and financial instruments in order to hedge short-term price movements. Higher oil prices may still lead a producer to invest in increased capacity.  Since lag times are significant in the offshore sector, we would then expect to see a multi-year lagged effect of prices on production.

However even with the question of lagged production and investment, some ambiguity exists.  \citet{mohn_efforts_2008} suggests and finds evidence for the idea that in periods of high oil prices offshore producers will invest more in risky wild-cat drilling in search of new fields. But when prices are low, they tend to concentrate investments in lower-risk ventures, like expanding production in existing fields. If this effect were to dominate, then it may even be plausible that a fall in prices leads to an \emph{increase} in production. Though the converse - an increase in prices leading to reduced production - is unlikely.

@@

For such a prominent subject, the lack of research on the role of price in oil field production is striking. Two main factors likely contribute to the limited literature - the availability of data and the non-linear time profile of field production.  Large private oil companies, notably the “super majors” and state-owned oil companies have historically accounted for the vast majority of oil production and reserves. \footnote{See for example the economist article titled Supermajordämmerung from August 3rd, 2013: \url{http://www.economist.com/news/briefing/21582522-day-huge-integrated-international-oil-company-drawing}} These entities tend to consider field-level data as either company or state secrets.

However, over the last few years, a movement towards making the petroleum and other extractive industries more transparent has taken form.  The Norwegian government has been on the forefront of this movement and increasingly committed itself to transparency in the petroleum sector. \footnote{See \url{http://www.regjeringen.no/en/sub/eiti---extractive-industries-tranparency/about-eiti.html?id=633586}} Over the last several years, detailed data on most aspects of the country's oil industry has become openly available.

In this article, I use historical production data from nearly all 77 currently or formerly oil-producing fields on the Norwegian Continental Shelf in order to estimate the effect that prices have on oil production.

By looking only at the effect of price on production in fields that currently or previously have produced oil, I am limiting the scope of this article.  The effect of oil prices on total production over the long term is due to both the reaction of production in existing fields as well as searching for new fields. This article is limited to the decisions oil companies can make to increase production in existing fields like making use of spare capacity or increasing production drilling. In fact, an implication of this work is that much of the total production response from higher oil prices is likely from increased searching as well as development of previously un-economic fields.

The main finding in this article is that oil production at the field level has no significant reaction to concurrent changes in the oil price.  A modest effect can be detected at a lag of between 2 and 4 years of between 2 to 7\% increase in yearly production for a 10 dollar increase in the price of oil.

@@

The main methodological problem is the non-linear production profile of oil fields.  Once full-scale extraction is started in an oil field, pressure in wells will start declining. Compensating measures such as gas and water injection have been successful, but their impact is temporary. In turn production rates will therefore inevitably decline.

More so, major oil field discoveries, starting times and in turn production are correlated across fields as the top panel of figure \ref{oil_decline} shows with the production profile of the 10 largest Norwegian oil fields.  The result is a total production curve that is bell-shaped over time as in the lower panel of figure \ref{oil_decline}.  Since the oil price series is autocorrelated and non-stationary, failure to properly account for the production profile will lead to spurious estimation of the price terms in a regression.

@@

The direction of this bias can be gleaned in figure \ref{oil_decline}.  High oil prices were present at periods of relatively low production in the late 1970s and early 1980s as well as over the last 10 years, however real prices reached some of their lowest levels at the same time as the top of production around the year 2000. These oil price dynamics almost certainly affected investment decisions in the industry and in turn total production \citep{osmundsen_is_2007}, \citep{aune_financial_2010}. However the inverse contemporaneous relationship at the field level is entirely coincidental, but will heavily bias the estimation of the effect of price on production if field production profiles are not properly accounted for.

@@

The results suggest that little to no concurrent effect of oil prices on production exists.  However I estimate modest effects on the first through fourth lagged terms. These results are in line with an understanding of the industry as being highly capital intensive with high operating costs.  Maintaining spare capacity in order to adjust to changing oil prices is likely prohibitively expensive.  Instead, production is increased through investment, which may take up to several years to show an effect.

\section{Econometric Literature on The Effect of Oil Price on Production}

@@

An extensive literature exists that attempts to estimate the effect of oil prices on regional or global oil supply. Early econometric attempts, like \citet{adelman_fea_1975} often made the assumption that the price contained all relevant information, ignoring the role of geophysical limits at the regional level. \citet{hamilton_oil_1983} showed that macroeconomic events could not fully explain movements in oil prices, rejecting the notion that prices contained all relevant information to oil supply. \citet{kaufmann_oil_1991} presents a relatively early attempt at combining both economic and geophysical aspects in an empirical model of oil production in the US. The author used a Hbbert-like \citep{hubbert_energy_1962} shape-fitting model to capture regional geophysical dynamics, using the residuals to model the effects of price on production. The author predicted that depletion and reduced oil supply from the lower 48 states would continue - a prediction that proved mostly accurate up until the technological revolution of fracking and shale oil extraction.

A similar analysis combining econometric models with shape-fitting models by \citet{cleveland_forecasting_1991} also led to pessimistic forecasts of future US production. Further refinements of these were done by, among others, \citet{pesaran_forecasting_1995}  and \citet{moroney_integrated_1999}. \citet{kaufmann_oil_2001} present an extension and generalisation using a Vector Error Correction Model (VECM) that relaxes the assumption that firms produce their fields in order of decreasing quality and also includes institutional factors in US oil production. They claim that they are able to account for most of the price variation between 1938 and 1991.

\citet{adelman_mineral_1990} and \citet{adelman_modelling_1993} takes an opposing view, using cointegration and causality testing to argue that scarcity, and in turn geophysical forces had little to do with the oil price movements of the 1970s and 1980s. Instead Adelman argues that cartel behavior of low cost producers likely stood behind increasing oil prices. \citet{lynch_forecasting_2002} provides a fuller overview of these competing strands of literature.

The economics literature on petroleum production tends to be overweighted towards studies of the US, but some global studies have also been attempted. \citet{watkins_world_1998} attempt to estimate supply functions for 41 countries by estimating the relationship between the price of undeveloped reserves and reserve additions over time. Based on the aggregated estimates, they argued that fears of increasing global scarcity at the time were overblown.

\citet{farzin_impact_2001} attempts to estimate the price elasticity of added global reserves and finds a small positive elasticity.  \citet{ramcharran_oil_2002} estimates a total supply function of oil from several OPEC countries based on data from 1973 to 1997.  The author finds a negative price elasticity for several of the countries, and interprets this as evidence of producers targeting revenue.  However since the author does not take into account the production profile of oil fields and the spurious correlation that can arise with autocorrelated prices, these estimates come under considerable doubt. \citet{ringlund_does_2008} estimates how rig activity reacts to changes in oil prices in non-Opec countries, and finds substantial heterogeneity based on the structure of a country's oil industry. Though in the long run, they find an average elasticity of close to one.

On the surface, this article fits well into the previous econometrics literature on petroleum supply in that there is a focus on establishing and decomposing the effects of geophysical and economic forces. However, most of the above articles focus on total long-term regional oil supply, which as I mention above, is the result of an extensive chain of decisions from surveying of a potential play, exploration drilling, to production. This article, on the other hand, is considerably narrower in scope - focusing exclusively on production in active fields. While ambiguity exists on whether geophysical forces and scarcity are significant factors that need to be controlled for in the long term regional perspective, they are without question dominant factors at the field level.

Microeconomic aspects of oil supply have more recently come to the fore. \citet{kellogg_effect_2014} studies the effect of oil-price uncertainty on drilling and exploration. The author finds that oil exploration firms in Texas approximately respond as real-option theory would predict when it comes to the timing of drilling.  A model and test using data from North Sea producers on the UK continental shelf by \citet{hurn_geology_1994}, on the other hand, fails to find evidence that the variance in the oil price affects the timing of oil field development.

I do not attempt to directly model uncertainty, however given that the investments needed to increase oil production in an existing field are to a certain extent irreversible and that oil prices are highly volatile, the results can and probably should be interpreted with the real options framework in mind.

Studies using detailed Norwegian data on offshore activity also exist, though the focus has mainly been on exploration and drilling.  \citet{mohn_exploration_2008} finds that long-term changes in the oil-price has a strong effect on exploratory drilling though little effect is measured from short-term changes in the oil price. \citet{osmundsen_exploration_2010} analyses drilling productivity over time on the Norwegian Continental Shelf while \citet{mohn_efforts_2008} finds that higher oil prices leads to higher reserves as well as that oil prices affect producer risk-preferences - with higher prices leading to lower success rates but larger discovery size. \citet{mohn_elastic_2010} presents a thorough econometric analysis of the role of economic and policy factors in oil exploration and reserve additions, as well as a useful summary of exploration and production on the Norwegian Continental Shelf.

Most of the microeconometric articles analyzing the petroleum industry, like the ones mentioned above, focus on issues of exploration, drilling, and rig activity. Microeconomic studies of production from operational fields are more sparse. \citet{black_is_1998} tests the relevance of “nesting” a structural empirical model of profit maximization that takes into account oil prices into a typical geo-engineering model of oil field production.  They find strong evidence that taking into account profit maximization - and implicitly price - substantially improves the fit compared to a purely geo-engineering model. The major limitation of their methodology is that they are only able to test whether including economic factors like price affects the fit of the model but are not able to give an estimate of the magnitude of the effect. Methodologically, the paper also relies heavily on strict assumptions about the functional form of both the geo-engineering and economic aspects of their model.  By taking a more flexible, semi-parametric approach to modeling oil field production profiles, this article avoids problems with overly restrictive assumptions.

\citet{rao_taxation_2010} uses data on onshore oil wells in California to estimate the effects of tax changes and price controls. The author finds that short-term tax changes caused small, but significant re-timing of production from oil wells.  These findings are generally consistent with the findings of this article, though the author finds a concurrent effect of changes in taxation on production while I do not find any concurrent effect of prices on production. The significant differences in cost and complexity between operating offshore and onshore likely account for the different results.  How producers react to a short-term tax change as opposed to a change in price also likely differs.

@@

The first commercial oil well in Norwegian continental waters was discovered in December of 1969 in what became the Ekofisk oil field. Most of the largest fields in the North Sea were found relatively early on while more recent finds have tended to be smaller - a pattern typical of oil producing areas called creaming. A major exception to this trend was the recent find of the Johan Sverdrup field which is estimated to have approximately 300 million Standard Cubic Meters (SM3) of recoverable oil. \footnote{The Johan Sverdrup field is estimated to begin producing oil in 2019 and is not present in the data set used for the analysis.}

Exploration in the Norwegian Sea was opened in the early 1980’s and the first commercial field started production in 1981.  While several mid-sized fields have been discovered, the Norwegian Sea has generally disappointed in terms of commercial oil finds.

Norwegian waters in the Barents Sea have also been open to exploration since the 1980's, however up until recently only a few, small finds were made and none came into commercial production.  Recently several significant oil and gas finds have been made in the Barents Sea - notably the oil field Goliat and the large gas find Sn\o hvit, which are both currently under development but not yet producing.  The agreement between Norway and Russia in June 2011 settling a long-running dispute over the maritime delimitation has also given a boost to new exploration in the region.

Profits from oil and gas production in Norway are subject to a resource tax of 51 \% on top of the ordinary income tax of 27 \%, thus income from petroleum production is taxed at a total marginal rate of 78 \%.  The central government also receives revenues through its majority ownership stake in Statoil. In addition, a wholly state-owned company, Petoro, manages the development licenses that the state retains - referred to as the State's Direct Financial Interest (SDFI).

The tax system is also designed to be neutral, that is a field that is profitable before tax should also be profitable after tax. This is accomplished by only taxing the net-profits of oil companies, allowing companies to deduct investment costs and deduct and carry forward losses.\footnote{The Norwegian Oil and Gas industry group Norsk Petroleum provides a fuller overview: \url{http://www.norskpetroleum.no/en/framework/petroleum-tax/}}

@@

More important to the offshore sector were accounting changes that were implemented in 2002 and 2005, which were meant to encourage new entrants by allowing losses to be carried forward for tax purposes and by introducing a rebate on the tax-value of losses associated with searching and drilling.  In general though, these rules mainly affected searching and discoveries of new fields rather than production from existing fields and I do not control for the tax changes in my model.  More information on taxation and revenues from the offshore sector can be found at the website of the Norwegian Ministry of Finance. \footnote{\url{http://www.regjeringen.no/en/dep/fin/Selected-topics/taxes-and-duties/bedriftsbeskatning/Taxation-of-petroleum-activities.html?id=417318}}

Rights to explore and eventually produce on the Norwegian Continental Shelf are based on a system where the government announces geographic blocks that will be opened to oil exploration and production subject to production licenses.  Production licenses are initially granted for between 4-6 years subject to requirements that firms are actively searching in their awarded blocks.  If oil or gas deposits are proven then the production license can be extended for up to 30 years.  In general, the frameworks are fair, predictable and stable for companies who find commercially extractable oil deposits and regulatory interference is unlikely to be the cause of any observed changes in production from existing oil fields.  For more information see the website of the Norwegian Petroleum Directorate. \footnote{\url{http://www.npd.no/en/Topics/Production-licences/Theme-articles/Production-licence--licence-to-explore-discover-and-produce-/}}

@@

\section{The Effect of Oil Price on Field Production}

>>>>

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{natbib} %for citet and citep
\usepackage{syntonly}
\usepackage{esdiff} %for writing partial derivatives
\usepackage{url} %for inserting urls
\usepackage{placeins} % for limitting floats
\usepackage{comment} %for excluding figures
\usepackage{booktabs} %for pretty tables
% \excludecomment{figure} % for printing without figures
% \syntaxonly % for quickly checking document
%set document settings

\doublespacing %from package setspacs

\title{The Effect of Oil Prices on Field Production: Evidence from the Norwegian Continental Shelf}
% \author{Johannes Mauritzen\\
		% Department of Economics\\
  %       BI Norwegian Business School\\
  %       Trondheim Campus\\
		% Trondheim, Norway\\
  %       johannes.mauritzen@bi.no\\
  %       \url{jmaurit.github.io}\\
		% }
\date{\today}

@@

\begin{abstract}
I use detailed field-level data on Norwegian offshore oil production and a semi-parametric additive model to control for the production profile of fields to estimate the effect of oil prices on production. I find no significant evidence of a concurrent reaction of field production to oil prices, though a modest lagged effect is found of the magnitude of approximately 2\% to 7\% for a \$10 per barrel increase in the real price of oil.\\
Keywords: Oil production, Oil Prices, Norway, Generalized Additive Model.
Words: 7100
\end{abstract}

@@

For most of the past century, crude oil has been the world's single most important and valuable fuel source.\footnote{Oil is the world's largest single energy source, consisting of approximately 33\% of total energy consumption in 2013. Oil is also the most valuable in terms of market price per unit of energy \citep{british_petroleum_statistical_2013}.} Naturally, how the oil price affects the world economy as well as how oil production reacts to the oil price have been fundamental topics in economics. Recent volatility in the price of oil has made this an especially relevant topic.

Oil and gas supply is the product of an often lengthy and complex chain of conditional investment decisions. Starting with the exploration phase, where a potential oil-producing region or ``play" is explored, leases are acquired, prospects are identified and exploration wells are drilled. In the case of a significant hydrocarbon find, a field is appraised to judge its size, difficulty of extraction and in turn profitability. For fields seen as sufficiently profitable, a development stage commences, where a detailed drilling and production plan is drawn up and submitted to obtain permits and regulatory approval. Operating agreements are signed and subcontractors acquired for offshore construction and well development. Finally, at the end of this multi-year process, a build-out can commence, where full production capacity is built out as fast as possible, followed by an extended depletion phase. All of this consists of what is referred to as ``upstream" activity, with steps such as shipping, refining and trading, among others, making up an equally complex ``downstream" process.\footnote{\citet{lefler_deepwater_2011} provides an accessible guide to the offshore petroleum exploration and production industry.}

A large body of economic literature exists on the complex chain of investment and production decisions made in the upstream oil and gas industry. This article looks at the important topic of the effect of oil prices on production from active offshore fields, where a significant gap exists in the literature.

While numerous studies have taken up the issue of how the price of oil affects searching for new fields as well as total oil production at both the regional and global level, few studies exist on the effects of oil prices on oil production at the field level. The studies that do use field- and well-level data, such as \citet{rao_taxation_2010}, often use data from onshore installations.  However, an increasing amount of the world's oil production comes from hard-to-reach offshore fields.  Production from challenging offshore environments is substantially different in character than from onshore installations.

The effect of oil prices on producing fields is an important topic in understanding the mechanisms of how total oil supply reacts to price. The response of oil production in existing fields is especially important now as many of the major oil-producing areas, like the Norwegian Continental Shelf, are mature and an increasing share of total investment will be directed at capacity enhancements in producing fields rather than new field developments. Figure \ref{investment} shows that investments in producing fields make up a large and growing share of total investment. How production from mature fields will respond to changes in oil prices has major implications for the oil industry, the state finances of oil-producing countries and long-run oil price formation.

\begin{figure}
	\center
	\includegraphics[width=.9\textwidth]{figures/investment.png}
	\caption{Investment (millions of NOK) in the oil and gas extraction industry on the Norwegian Continental Shelf. A large and growing share of total investment goes into fields that are already producing as opposed to investment in search, exploration and build-out of new fields.}
	\label{investment}
\end{figure}

The question of the effect of oil prices on production and the more general question of optimal oil extraction has spawned a large theoretical literature dating back to the seminal work of \citet{hotelling_economics_1931}. \citet{krautkraemer_nonrenewable_1998} provides a good overview. At a basic level a central idea of much of this theoretical work is that with a non-renewable resource, production involves a significant opportunity cost: more production in the current period means less production in future periods.  Within these frameworks, prices and expectations of prices become important variables in the production decision. A simple Hotelling model suggests that a producer would immediately change its production in response to a change in oil prices in order to dynamically optimize the total extraction.

But in practice, the question is not as clear cut.  Production on the Norwegian Continental Shelf -- as in most other offshore production areas -- has extremely high fixed and operating costs.  Keeping spare capacity available in order to adjust to changing oil prices may simply be too expensive. Producers may instead use storage and financial instruments in order to hedge short-term price movements. Higher oil prices may still lead a producer to invest in increased capacity and production drilling. However, since lag times are significant in the offshore sector, we would then expect to see a multi-year lagged effect of prices on production.

However, even with the question of lagged production and investment, some ambiguity exists.  \citet{mohn_efforts_2008} finds evidence for the idea that in periods of high oil prices offshore producers will invest more in risky wild-cat drilling in search of new fields. But when prices are low, they tend to concentrate investments in lower-risk ventures, like expanding production in existing fields. If this effect were to dominate, then it may even be plausible that a fall in prices leads to an \emph{increase} in production. Though the converse, an increase in prices leading to reduced production, is unlikely.

@@

For such a prominent subject, the lack of research on the role of prices in oil field production is striking. Two main factors likely contribute to the limited literature: The availability of data and the non-linear time profile of field production.  Large private oil companies, notably the ``super majors" and state-owned oil companies, have historically accounted for the vast majority of oil production and reserves.\footnote{See for example the economist article titled Supermajordämmerung from August 3rd, 2013: \url{http://www.economist.com/news/briefing/21582522-day-huge-integrated-international-oil-company-drawing}} These entities tend to consider field-level data as either company or state secrets.

However, over the last few years, a movement towards making the petroleum and other extractive industries more transparent has taken form.  The Norwegian government has been on the forefront of this movement and has increasingly committed itself to transparency in the petroleum sector.\footnote{See \url{http://www.regjeringen.no/en/sub/eiti---extractive-industries-tranparency/about-eiti.html?id=633586}} Over the last several years, detailed data on most aspects of the country's oil industry have become openly available.

In this article, I use historical production data from nearly all 77 currently or formerly oil-producing fields on the Norwegian Continental Shelf in order to estimate the effect that oil prices have on oil production.

By looking only at the effect of oil prices on production in fields that currently or previously have produced oil, I am limiting the scope of this article.  The effect of oil prices on total production over the long term is due not only to how production reacts in existing fields, but also to changes in search activity for new fields. This article is limited to the decisions oil companies make to increase production in existing fields like making use of spare capacity and increasing production drilling. In fact, an implication of this work is that much of the total production response to higher oil prices is likely from increased searching as well as from development of previously un-economic fields.

The main finding in this article is that oil production at the field level has no significant reaction to concurrent changes in the oil price.  A modest effect can be detected at a lag of between 2 and 4 years of between 2\% to 7\% increase in yearly production for a \$10 increase in the price of oil.

@@

The main methodological problem in the estimation is the non-linear production profile of oil fields. Once full-scale extraction is started in an oil field, pressure in wells will start declining. Compensating measures such as gas and water injection have been successful, but their impact is temporary. Production rates will therefore inevitably decline.

Moreover, major oil field discoveries, starting times, and in turn production are correlated across fields. The top panel of figure \ref{oil_decline} shows the production profile of the 10 largest Norwegian oil fields. The temporal correlation between the fields results in a total production curve that is bell-shaped over time as in the lower panel of figure \ref{oil_decline}.  Since the oil price series is autocorrelated and non-stationary, failure to properly account for the production profile will lead to spurious estimation of the price terms in a regression.

@@

The direction of this bias can be gleaned in figure \ref{oil_decline}.  High oil prices were present at periods of relatively low production in the late 1970s and early 1980s as well as over the last 10 years. However, real prices reached some of their lowest levels at the same time as the peak of production around the year 2000. These oil price dynamics almost certainly affected investment decisions in the industry and in turn total production \citep{osmundsen_is_2007,aune_financial_2010}. However, the inverse contemporaneous relationship at the field level is entirely coincidental. This will heavily bias the estimates of the effect of oil prices on production if field production profiles are not properly accounted for.

@@

The results suggest that little to no concurrent effect of oil prices on production exists.  However, I estimate modest effects on the second through fourth yearly lagged terms. These results are in line with an understanding of the industry as being highly capital intensive with high operating costs.  Maintaining spare capacity in order to adjust to changing oil prices may be prohibitively expensive.  Instead, production adjusts only with a lag, the result of a change in investment.

\section{Econometric Literature on the Effect of Oil Prices on Production}

@@

An extensive literature exists that attempts to estimate the effect of oil prices on regional or global oil supply. Early econometric studies, such as \citet{adelman_fea_1975}, often made the assumption that the price contained all relevant information, ignoring the role of geophysical limits at the regional level. \citet{hamilton_oil_1983} showed that macroeconomic events could not fully explain movements in oil prices, rejecting the notion that prices contained all information relevant to oil supply. \citet{kaufmann_oil_1991} presents a relatively early attempt at combining both economic and geophysical aspects in an empirical model of oil production in the US. The author used a Hubbert-like \citep{hubbert_energy_1962} shape-fitting model to capture regional geophysical dynamics, using the residuals to model the effects of price on production. The author predicted that depletion and reduced oil supply from the lower 48 states would continue -- a prediction that proved mostly accurate up until the technological revolution of fracking and shale oil extraction.

A similar analysis combining econometric models with shape-fitting models by \citet{cleveland_forecasting_1991} also led to pessimistic forecasts of future US production. \citet{pesaran_forecasting_1995} and \citet{moroney_integrated_1999}, among others, made further refinements of these types of models. \citet{kaufmann_oil_2001} present an extension and generalization using a Vector Error Correction Model (VECM) that relaxes the assumption that firms produce their fields in order of decreasing quality and also includes institutional factors in US oil production. They claim that they are able to account for most of the price variation between 1938 and 1991.

\citet{adelman_mineral_1990, adelman_modelling_1993} takes an opposing view, using cointegration and causality testing to argue that scarcity, and in turn geophysical forces had little to do with the oil price movements of the 1970s and 1980s. Instead Adelman argues that the cartel behavior of low cost producers likely stood behind increasing oil prices. \citet{lynch_forecasting_2002} provides a fuller overview of these competing strands of literature.

The economics literature on petroleum production tends to be overweighted towards studies of the US, but some global studies have also been attempted. \citet{watkins_world_1998} attempt to estimate the supply functions for 41 countries by modeling the relationship between the price of undeveloped reserves and reserve additions over time. Based on the aggregated estimates, they argued that fears of increasing global scarcity at the time were overblown. \citet{farzin_impact_2001} attempts to estimate the price elasticity of added global reserves and finds a small positive elasticity.  \citet{ramcharran_oil_2002} estimates a total supply function of oil from several OPEC countries based on data from 1973 to 1997.  The author finds a negative price elasticity for several of the countries, and interprets this as evidence of producers targeting revenue.  However since the author does not take into account the production profile of oil fields and the spurious correlation that can arise with autocorrelated prices, these estimates come under some doubt. \citet{ringlund_does_2008} estimates how rig activity reacts to changes in oil prices in non-OPEC countries, and finds substantial heterogeneity based on the structure of a country's oil industry. Though in the long run, they find an average elasticity of close to one.

This article fits well into the previous econometrics literature on petroleum supply in that there is a focus on establishing and decomposing the effects of geophysical and economic forces. However, most of the above articles focus on total long-term regional oil supply, which, as I mention above, is the result of an extensive chain of decisions from surveying of a potential play, exploration drilling, to production. This article, on the other hand, is considerably narrower in scope -- focusing exclusively on production in active fields. While ambiguity exists on whether geophysical forces and scarcity are significant factors that need to be controlled for in the long term regional perspective, they are without question dominant factors at the field level.

Microeconomic aspects of oil supply have more recently come to the fore. \citet{kellogg_effect_2014} studies the effect of oil-price uncertainty on drilling and exploration. The author finds that oil exploration firms in Texas approximately respond as real-option theory would predict when it comes to the timing of drilling.  A model and test using data from North Sea producers on the UK continental shelf by \citet{hurn_geology_1994}, on the other hand, fails to find evidence that the variance in oil prices affects the timing of oil field development. I do not attempt to directly model uncertainty, however given that the investments needed to increase oil production in an existing field are to a certain extent irreversible and that oil prices are highly volatile, the results can and probably should be interpreted with the real options framework in mind.

Studies using detailed Norwegian data on offshore activity also exist, though the focus has mainly been on exploration and drilling.  \citet{mohn_exploration_2008} finds that long-term changes in the oil-price has a strong effect on exploratory drilling though little effect is measured from short-term changes in the oil price. \citet{osmundsen_exploration_2010} analyses drilling productivity over time on the Norwegian Continental Shelf while \citet{mohn_efforts_2008} finds that higher oil prices leads to higher reserves as well as that oil prices affect producer risk-preferences -- with higher prices leading to lower success rates but larger discovery size. \citet{mohn_elastic_2010} presents a thorough econometric analysis of the role of economic and policy factors in oil exploration and reserve additions, as well as a useful summary of exploration and production on the Norwegian Continental Shelf.

Most of the microeconometric studies analyzing the petroleum industry, like the ones mentioned above, focus on issues of exploration, drilling, and rig activity. Microeconomic studies of production from operational fields are more sparse. \citet{black_is_1998} tests the relevance of “nesting” a structural empirical model of profit maximization that takes into account oil prices into a typical geo-engineering model of oil field production.  They find strong evidence that taking into account profit maximization -- and implicitly oil prices -- substantially improves the fit compared to a purely geo-engineering model. The major limitation of their methodology is that they are only able to test whether including economic factors like oil prices affects the fit of the model but are not able to give an estimate of the magnitude of the effect. Methodologically, the paper also relies heavily on strict assumptions about the functional form of both the geo-engineering and economic aspects of their model.  By taking a more flexible, semi-parametric approach to modeling oil field production profiles, this article avoids problems with overly restrictive assumptions.

\citet{rao_taxation_2010} uses data on onshore oil wells in California to estimate the effects of tax changes and price controls. The author finds that short-term tax changes caused small, but significant re-timing of production from oil wells.  These findings are generally consistent with the findings of this article, though the author finds a concurrent effect of changes in taxation on production while I do not find any concurrent effect of prices on production. The significant differences in cost and complexity between operating offshore and onshore likely account for the different results.  How producers react to a short-term tax change as opposed to a change in the oil price also likely differs.

@@

The first commercial oil well on the Norwegian Continental Shelf was discovered in December of 1969 in what became the Ekofisk oil field. Most of the largest fields in the North Sea were found relatively early on while more recent finds have tended to be smaller -- a pattern typical of oil producing area called ``creaming''. A major exception to this trend was the recent find of the Johan Sverdrup field which is estimated to have approximately 300 million Standard Cubic Meters (SM3) of recoverable oil.\footnote{The Johan Sverdrup field is estimated to begin producing oil in 2019 and is not present in the data set used for the analysis.}

Exploration in the Norwegian Sea was opened in the early 1980s and the first commercial field started production in 1981.  While several mid-sized fields have been discovered, the Norwegian Sea has generally disappointed in terms of commercial oil finds.

Norwegian waters in the Barents Sea have also been open to exploration since the 1980s. However, up until recently only a few, small finds were made and none came into commercial production.  Recently, several significant oil and gas finds have been made in the Barents Sea -- notably the oil field Goliat and the large gas find Sn\o hvit, which are both currently under development but not yet producing.  The agreement between Norway and Russia in June 2011 settling a long-running dispute over the maritime delimitation has also given a boost to new exploration in the region.

Profits from oil and gas production in Norway are subject to a resource tax of 51\% on top of the ordinary income tax of 27\%, thus income from petroleum production is taxed at a total marginal rate of 78\%.  The State also receives revenues through its majority ownership stake in Statoil. In addition, a wholly state-owned company, Petoro, manages the development licenses that the State retains -- referred to as the ``State's Direct Financial Interest (SDFI).''

The tax system is also designed to be neutral. That is, a field that is profitable before tax should also be profitable after tax. This is accomplished by only taxing the net-profits of oil companies, allowing companies to deduct investment costs and deduct and carry forward losses.\footnote{The Norwegian Oil and Gas industry group Norsk Petroleum provides a fuller overview: \url{http://www.norskpetroleum.no/en/framework/petroleum-tax/}}

@@

More important to the offshore sector were accounting changes that were implemented in 2002 and 2005, which were meant to encourage new entrants by allowing losses to be carried forward for tax purposes and by introducing a rebate on the tax-value of losses associated with searching and drilling.  In general though, these rules mainly affected searching and discoveries of new fields rather than production from existing fields and I do not control for the tax changes in my model.  More information on taxation and revenues from the offshore sector can be found at the website of the Norwegian Ministry of Finance.\footnote{\url{http://www.regjeringen.no/en/dep/fin/Selected-topics/taxes-and-duties/bedriftsbeskatning/Taxation-of-petroleum-activities.html?id=417318}}

Rights to explore and eventually produce on the Norwegian Continental Shelf are based on a system where the government announces geographic blocks that will be opened to oil exploration and production subject to production licenses.  Production licenses are initially granted for between 4-6 years subject to requirements that firms are actively searching in their awarded blocks.  If oil or gas deposits are proven then the production license can be extended for up to 30 years.  In general, the frameworks are fair, predictable and stable for companies who find commercially extractable oil deposits and regulatory interference is unlikely to be the cause of any observed changes in production from existing oil fields.  For more information see the website of the Norwegian Petroleum Directorate.\footnote{\url{http://www.npd.no/en/Topics/Production-licences/Theme-articles/Production-licence--licence-to-explore-discover-and-produce-/}}

@@

\section{The Effect of Oil Prices on Field Production}
